---
title: "Monthly Finance Report"
subtitle: "`r format(as.Date(paste0(params$month, '-01')), '%B %Y')`"
author: "Finance Tracker"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-float: true
    theme: cosmo
    embed-resources: true
  pdf:
    toc: true
    geometry: margin=2.5cm
    colorlinks: true
params:
  month: "2026-04-01"
  data_path: "data/sample_expenses.csv"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(lubridate)
library(readr)
library(ggplot2)
library(scales)
library(knitr)

source("R/data_pipeline.R")

df        <- load_csv(params$data_path)
month_lbl <- format(as.Date(params$month), "%B %Y")
total     <- sum(df$amount, na.rm = TRUE)
n_trans   <- nrow(df)
top_cat   <- df %>% group_by(category) %>% summarise(t=sum(amount)) %>% slice_max(t,n=1)
```

## Summary

This report covers all recorded expenses for **`r month_lbl`**.

| Metric | Value |
|--------|-------|
| Total spent | `r paste0("€", format(round(total,2), big.mark=","))` |
| Number of transactions | `r n_trans` |
| Average per transaction | `r paste0("€", round(total/max(n_trans,1),2))` |
| Top spending category | `r top_cat$category` (`r paste0("€", round(top_cat$t,2))`) |

---

## Spending by category

```{r cat-chart, fig.height=4, fig.width=7}
cat_df <- category_summary(df)

ggplot(cat_df, aes(x = reorder(category, total), y = total, fill = category)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = paste0("€", round(total, 0))),
            hjust = -0.1, size = 3, color = "#444") +
  coord_flip() +
  scale_fill_manual(values = CATEGORY_COLORS) +
  scale_y_continuous(labels = dollar_format(prefix="€"), expand = expansion(mult=c(0,.2))) +
  labs(x = NULL, y = NULL, title = paste("Category breakdown —", month_lbl)) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(size=13, face="bold"))
```

```{r cat-table}
cat_df %>%
  mutate(
    total     = paste0("€", formatC(total, format="f", digits=2, big.mark=",")),
    avg_trans = paste0("€", round(avg_trans, 2))
  ) %>%
  rename(Category=category, `Total spent`=total,
         Transactions=n_trans, `Avg per transaction`=avg_trans) %>%
  kable(align=c("l","r","r","r"))
```

---

## Daily spending timeline

```{r daily-chart, fig.height=3.5, fig.width=8}
daily <- df %>%
  group_by(date) %>%
  summarise(total = sum(amount, na.rm=TRUE), .groups="drop")

ggplot(daily, aes(x=date, y=total)) +
  geom_col(fill="#1D9E75", alpha=0.8, width=0.7) +
  scale_x_date(date_breaks="1 week", date_labels="%d %b") +
  scale_y_continuous(labels=dollar_format(prefix="€")) +
  labs(x=NULL, y=NULL, title="Daily spend") +
  theme_minimal(base_size=11) +
  theme(panel.grid.minor=element_blank())
```

---

## Budget vs actual

```{r budget-table}
budget_status(df, DEFAULT_BUDGETS) %>%
  mutate(
    budget    = paste0("€", formatC(budget, format="f", digits=0, big.mark=",")),
    total     = paste0("€", formatC(total,  format="f", digits=2, big.mark=",")),
    remaining = paste0("€", formatC(as.numeric(gsub("[^0-9.-]","",budget)) - 
                                     as.numeric(gsub("[^0-9.-]","",total)),
                                    format="f", digits=2)),
    status    = toupper(status)
  ) %>%
  select(category, budget, total, pct, status) %>%
  rename(Category=category, Budget=budget, Spent=total, `% used`=pct, Status=status) %>%
  kable(align=c("l","r","r","r","c"))
```

---

## Top merchants

```{r merchant-table}
top_merchants(df, 10) %>%
  mutate(total = paste0("€", formatC(total, format="f", digits=2, big.mark=","))) %>%
  rename(Merchant=merchant, `Total spent`=total, Visits=n) %>%
  kable(align=c("l","r","r"))
```

---

## All transactions

```{r all-transactions}
df %>%
  arrange(desc(date)) %>%
  select(date, category, merchant, description, amount) %>%
  mutate(amount = paste0("€", formatC(amount, format="f", digits=2))) %>%
  rename(Date=date, Category=category, Merchant=merchant,
         Description=description, Amount=amount) %>%
  kable(align=c("l","l","l","l","r"))
```

---

*Generated by Finance Tracker · `r Sys.time()`*
